---
title: "<b>Week 5: Example & Write Up </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "TOM BOOTH & ALEX DOUMAS"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

---
# Week's Learning Objectives

1. Understand the difference between outliers and influential points. 

2. Understand and apply methods to detect outliers and influential points.

3. Comment on the issues of correlation vs causation and generalisability of the sample results.

4. Run a full simple regression analysis from start to finish and to correctly interpret the results.

---
# Topics for today
+ Run an example analysis

  + Start to finish

  + In R
  
  + Live!

+ Check LEARN for the dates (will be recorded)

---
# A comment on examples

+ From this point on we will try to make use of published studies as examples in lectures.

+ These studies will be selected because they...
--
  + **are in open source journals** so we can read it
--
  + **have openly available data** so we can analyse it
--
  + **have openly available code** so we can see exactly how the authors did it
--

+ The use of an example does not mean we...
--
  + **agree with the authors**; academia is a place for debate, we won't censor
--
  + **think they are perfect studies**; no study is perfect and careful critique is good practice
--
  + **think they are perfect write ups**; there are many styles with pros and cons


---
# Next tasks

+ This week:
  + Complete your lab
  + Come to office hours
  + Weekly quiz: Assessed quiz - Week 4 content.
      + Open Monday 09:00
      + Closes Sunday 17:00