---
title: "<b>Week 4: Binary Predictors </b>"
subtitle: "Data Analysis for Psychology in R 2<br><br> "
author: "TOM BOOTH & ALEX DOUMAS"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
library(effsize)

theme_set(theme_gray(base_size = 15))

knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.retina = 1.5)
```

# Weeks Learning Objectives
1. Understand the meaning of model coefficients in the case of a binary predictor.

2. Be able to state the assumptions underlying a linear model.

3. Understand how to assess if a fitted model satisfies the linear model assumptions.

4. Understand how to use transformations when the model violates assumptions.


---
# Topics for today
+ Today will will focus on the linear model with a binary predictor.

  + Recap categorical and binary variables

  + Extend our data example

  + Run through the steps and interpretation for the linear model

---
#  Recap: Categorical variables 
+ Categorical variables can only take discrete values
	+ E.g., animal type: 1= duck, 2= cow, 3= wasp

+ They are mutually exclusive
  + No duck-wasps or cow-ducks!

+ In R, categorical variables should be of class `factor`
  + The discrete values are `levels`
  + Levels can have numeric values (1, 2 3) and labels (e.g. "duck", "cow", "wasp")

---
# Recap: Binary variable
+ Binary variable is a categorical variable with two levels.

+ Traditionally coded with a 0 and 1
  + Referred to as dummy coding
  + We will come back to this for categorical variables with 2+ levels

--

+ Why 0 and 1?
  + Quick version: It has some nice properties when it comes to interpretation.


---
# Extending our example

.pull-left[
+ Our in class example so far has used test scores and revision time for 10 students.

+ Let's say we collect this data on 150 students.

+ We also collected data on who they studied with;
  + 0 = alone
  + 1 = with others
  
+ So our variable `study` is a binary
]

.pull-right[

+ This data set is available on LEARN

```{r, warning=FALSE, message=FALSE}
df <- read_csv("./dapr2_lec07.csv") #<<
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(df)%>%
  kable_styling(., full_width = F) %>%
  scroll_box(height = "300px")
```

]

---
#  LM with binary predictors 
+ In previous lectures we asked the question:

  + **Do students who study more get higher scores on the test?**

+ And we specified a linear model:

$$y_i = \beta_0 + \beta_1 x_{i} + \epsilon_i$$

+ Or

$$score_i = \beta_0 + \beta_1 hours_{i} + \epsilon_i$$

--

+ And nothing changes with our binary variable. We can ask the question:

  + **Do students who study with others score better than students who study alone?**

$$score_i = \beta_0 + \beta_1 study_{i} + \epsilon_i$$

---
# In `R`

```{r}
res <- lm(score ~ study, data = df)
summary(res)
```


---
# Interpretation

.pull-left[
+ As before, the intercept $\hat \beta_0$ is the expected value of $y$ when $x=0$

+ What is $x=0$ here?
  + It is the students who study alone.

+ So what about $\hat \beta_1$?

+ **Look at the output on the right hand side.** 
  + What do you notice about the difference in averages?

]

.pull-right[
```{r warning=FALSE, message=FALSE}
df %>%
  group_by(., study) %>% #<<
  summarise(
    Average = round(mean(score),4) #<<
  )
```


]


---
# Interpretation
+ $\hat \beta_0$ = predicted expected value of $y$ when $x = 0$
  + Or, the mean of group coded 0 (those who study alone)
  
+ $\hat \beta_1$ = predicted difference between the means of the two groups.
  + Group 1 - Group 0 (Mean `score` for those who study with others - mean `score` of those who study alone)
  
+ Notice how this maps to our question. 
  + Do students who study with others score better than students who study alone?
  

---
class: center, middle
# Time for a break

**Have a go at the binary variable quiz**

---
class: center, middle
# Welcome Back!

**Where we left off... **

Let's think about the interpretation by group a little more.

---
#  Equations for each group 
+ What would our linear model look like if we added the values for $x$.


$$\widehat{score} = \hat \beta_0 + \hat \beta_1 study$$


+ For those who study alone ( $study = 0$ ):


$$\widehat{score}_{alone} = \hat \beta_0 + \hat \beta_1 \times 0$$


+ So;


$$\widehat{score}_{alone} = \hat \beta_0$$



---
#  Equations for each group 
+ For those who study with others ( $study = 1$ ):


$$\widehat{score}_{others} = \hat \beta_0 + \hat \beta_1 \times 1$$


+ So;


$$\widehat{score}_{others} = \hat \beta_0 + \hat \beta_1$$


+ And if we re-arrange;


$$\hat \beta_1 = \widehat{score}_{others} - \hat \beta_0$$


+ Remembering that $\widehat{score}_{alone} = \hat \beta_0$, we finally obtain:

$$\hat \beta_1 = \widehat{score}_{others} - \widehat{score}_{alone}$$


---
#  Visualize the model

```{r, echo=FALSE}
bin <- df %>%
  ggplot(., aes(x=factor(study), y=score, colour = study)) +
  geom_point(alpha=0.4) +
  labs(x = "\n Study", y = "Test Score \n") +
  ylim(0,10) +
  scale_x_discrete(labels = c("alone", "others")) +
  theme(legend.position = "none")

bin
```


---
#  Visualize the model
```{r, echo=FALSE}
bin +
  geom_jitter(width = .1, height = 0, alpha=0.4)

```


---
#  Visualize the model
```{r, echo=FALSE}
gpM <- df %>%
  group_by(study) %>%
  summarise(
    score = mean(score)
  )

bin +
  geom_jitter(width = .1, height = 0, alpha=0.4) +
  geom_errorbar(data = gpM, width=0.6,aes(ymax=..y..,ymin=..y..), size=1)

```



---
#  Visualize the model

```{r, echo=FALSE}
library(latex2exp)

bin +
  geom_jitter(width = .1, height = 0, alpha=0.4) +
  geom_errorbar(data = gpM, width=0.6, aes(ymax=..y..,ymin=..y..), size=1)+
  geom_segment(x=1.5, y=gpM[[1,2]], xend=1.5, yend=gpM[[2,2]], size =1, col="red") +
  geom_segment(x=1.48, y=gpM[[1,2]], xend=1.52, yend=gpM[[1,2]], size =1, col="red") +
  geom_segment(x=1.48, y=gpM[[2,2]], xend=1.52, yend=gpM[[2,2]], size =1, col="red") +
  geom_text(x=1.55, y = 5.5, label = TeX('$\\hat{\\beta}_1$'), size=5, col = "red") +
  geom_text(x=0.65, y = gpM[[1,2]] , label = TeX('$\\hat{\\beta}_0$'), size=5, col = "red")

```


---
#  Evaluation of model and significance of $\beta_1$

+ $R^2$ and $F$-ratio interpretation are identical to their interpretation in models with only continuous predictors.

+ And we assess the significance of predictors in the same way

+ We use the standard error of the coefficient to construct:
  + We calculate the $\hat \beta_1$ = difference between groups
	+ $t$-value and associated $p$-value for the coefficient
	+ Or a confidence interval around the coefficient


---
# Hold on... it's a t-test

```{r}
df %>%
  t.test(score ~ study, .)
```

???
Yup!


---
# Standardizing $\hat \beta_1$

+ When discussing continuous predictors we discussed standardized $b_1$ and unstandardized $\hat \beta_1$

+ Recall, when we calculated $b_1$ we used the SD of x, $s_x$.

+ For a binary categorical variable, the SD is not appropriate.

+ 1 unit also has meaning - it is the membership of a different group.

+ As such, we do not standardize


---
# Summary of today

+ Recapped categorical and binary variables

+ Introduced the linear model with a single binary variable

+ Considered the interpretation of the coefficients
  + And saw it is an independent sample $t$-test

---
# Next tasks

+ This week:
  + Complete your lab
  + Come to office hours
  + Weekly quiz: Assessed quiz - Week 3 content.
      + Open Monday 09:00
      + Closes Sunday 17:00